\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-17 doc/mlsys_paper/paper.tex
Slight reworking of sections and add plotting experiment.
It's a bit too long now, and there are still a few references to add, but this
took way longer than expected and I need to go to bed...

In this framework, if a user wants to optimize a differentiable objective
function, they only need to provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that {\tt ensmallen}
provides.  Table~\ref{tab:functionality} contrasts 
the classes of objective functions that can be handled by {\tt ensmallen}
and other popular frameworks and libraries.

@@

The required API for separable differentiable objective functions (i.e.~those
that would use an optimizer like SGD) is very similar, except that
\texttt{Evaluate()}, \texttt{Gradient()} and \texttt{EvaluateWithGradient()}
should operate only on mini-batches.  The same pattern applies for other
types of objective functions: only a few methods specific to class of objective
function itself must be implemented and then any optimizer may be used.

@@

\vspace*{-0.3em}
\section{Available Optimizers}
\vspace*{-0.5em}

Due to the straightforward abstraction framework, {\tt ensmallen} is able to
provide a large set of diverse optimization algorithms.  Below is a list of
currently available optimizers:

@@

However, both Python and GNU Octave have routes for acceleration, such as
numba~\cite{TODO} and Octave's built-in JIT.  Therefore, we hand-optimized the
Rosenbrock implementation using numba and ran the Octave implementation with the
JIT.  For the numba implementation, the underlying {\tt anneal.anneal()}
function had to be significantly modified.  These did produce some acceleration,
as can be seen in the second row of Table~\ref{tab:rosenbrock_results}.

@@

>>>>

Due to its straightforward abstraction framework, {\tt ensmallen} is able to
provide a large set of diverse optimization algorithms for objective functions
with these properties.  Below is a list of currently available optimizers:

In {\tt ensmallen}'s framework, if a user wants to optimize a differentiable objective
function, they only need to provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that {\tt ensmallen}
provides.  Table~\ref{tab:functionality} contrasts 
the classes of objective functions that can be handled by {\tt ensmallen}
and other popular frameworks and libraries.

@@

The required API for separable differentiable objective functions (i.e.~those
that would use an optimizer like SGD) is very similar, except that
\texttt{Evaluate()}, \texttt{Gradient()} and \texttt{EvaluateWithGradient()}
should operate only on mini-batches, and utility methods {\tt Shuffle()} and
{\tt NumFunctions()} must be added.  The same pattern applies for other
types of objective functions: only a few methods specific to class of objective
function itself must be implemented and then any optimizer may be used.

@@

@@

However, both Python and GNU Octave have routes for acceleration, such as
numba~\cite{TODO} and MEX bindings for Octave/MATLAB.  Therefore, we hand-optimized the
Rosenbrock implementation using numba and ran the Octave implementation with the
JIT.  For the numba implementation, the underlying {\tt anneal.anneal()}
function had to be significantly modified.  These did produce some acceleration,
as can be seen in the second row of Table~\ref{tab:rosenbrock_results}.  For
Octave, a MEX binding did not produce a noticeable difference.

@@

Lastly, we show the simplicity of using different optimizers for the same task.
Using a version of {\tt LinearRegressionFunction} adapted for separable
differentiable optimizers, we run six optimizers with default parameters in
eight lines of code:

% RC: I know this is inline code... but does it work here?  If not we can put it
% in a figure.
\begin{minted}[fontsize=\footnotesize]{c++}
  LinearRegressionFunction lrf(X, y); // X and y are data

  ens::StandardSGD<>().Optimize(lrf, sgdModel);
  ens::Adam().Optimize(lrf, adamModel);
  ens::AdaGrad().Optimize(lrf, adagradModel);
  ens::SMORMS3().Optimize(lrf, smorms3Model);
  ens::SPALeRASGD().Optimize(lrf, spaleraModel);
  ens::RMSProp().Optimize(lrf, rmspropModel);
\end{minted}

This pluggability makes it easy to use {\tt ensmallen} to generate comparisons
between optimizers, and in Figure \ref{fig:learning_curve} we use these six
lines of code to generate learning curves for each of these six optimizers on
the {\tt YearPredictionMSD} dataset from the UCI dataset
repository~\cite{ucimlrepository}. % TODO: add to bibtex
In addition, any other optimizers for separable differentiable objective
functions could be dropped into place.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{experiments/learning_curves.png}
\label{fig:learning_curve}
\caption{Learning curves for six {\tt ensmallen} optimizers on {\tt
YearPredictionMSD} dataset over 5 epochs of training.  Optimizers can be
tuned for better performance.}
\end{figure}
